---
title: "Chapter 02"
output: html_document
---

# Data
```{r data}
library(ggplot2)
set.seed(1)
n = 100
D = data.frame(
  x = rnorm(n),
  y = rnorm(n)
)
```

# 2.1 Covariance

Deviation from Mean and Product of Deviations from the Mean:

```{r Deviaion}
D$devp = (D$x-mean(D$x)) * (D$y-mean(D$y))
sum(D$devp)/(n-1)
cov(D$x, D$y)
```

Covariance is the sum of the product of x and y's deviation from the mean divided by n-1.

```{r Covariance}
D$devp.col = ifelse(D$devp<=0, "black", "red")
ggplot(D, aes(x,y)) +
  geom_point(color=D$devp.col)
```

Sign of the product of the deviation of the mean roughly equal to the *direction* of the linear relationship but nothing
about the *strength* of that relationship.

# 2.2 Correlation Coefficient

Standardize both X and Y:

```{r Standardize}
D$zx = (D$x-mean(D$x))/sd(D$x)
D$zy = (D$y-mean(D$y))/sd(D$y)
(sum(D$zx*D$zy))/(n-1)
cor(D$x, D$y)
```

Measures both the *direction* and *strength* of the linear relationship between variables.

# 2.4 Simple Linear Regression

> How do you know, in advance, if the relationship between X and Y is indeed
> ameanable to modeling with a linear equation?